---
title: "LHPRISM_Data_Analysis"
author: "Alexandra Tejeda"
date: "2023-08-21"
output:
  word_document: default
  html_document: default
  pdf_document: default
---e
---

# Inputting dataset and required packages.

```{r}
library(dplyr)
library(knitr)
library(tidyverse)
library(readr)
lhprism <- read_csv("LHPRISM%2DUpdated%2Dcopy%2Dcsv%2Ecsv-6.txt")
```

#### Fixed inconsistencies in OpenRefine.

-   Trimmed white space.
-   Removed commas from large numerical values.
-   Changed columns from categorical to numeric where appropriate.

#### Using separate function to split SITE CODE column into 3 separate columns.

```{r}
# Making new column for number of visits for each Site Code.
lhprism <- lhprism %>%
  group_by(`Site Code`) %>%
  mutate(Tot_Visits = length(unique(VISIT_DATE)))

lhprism <- lhprism %>% 
  separate(`Site Code`, c("Property_Code","Species_Code","Pop_Number"))

lhprism$Property_Pop <- paste(lhprism$Property_Code, lhprism$Pop_Number, sep = "_")
```

#### Utilized the lubridate package to replace the original VISIT_DATE column with a year-month-date format.

```{r}
lhprism$VISIT_DATE <- ymd(paste(lhprism$YEAR, lhprism$MONTH, lhprism$DAY, sep = "-"))

summary(lhprism$VISIT_DATE)
```

#### Finding the date that previously failed to parse.

```{r}
which(lhprism$VISIT_DATE == FALSE)

missing_dates <- lhprism[is.na(lhprism$VISIT_DATE), c("YEAR", "MONTH", "DAY")]
head(missing_dates)
```

# Making summary statistics tables.

```{r}
#NaN values occurred from 2013 to 2019 because there are no recorded cases of SEARCH_SQFT values until 2020. 

lhprism_yearly_stat <- lhprism %>%
    filter(!is.na(SEARCH_SQFT) | YEAR >= 2013) %>%
  group_by(YEAR) %>%
  summarize(across(c("SEARCH_SQFT"), ~ mean(.x, na.rm = TRUE))) 
            
kable(lhprism_yearly_stat, col.names = gsub("[_]", " ", names(lhprism_yearly_stat)), align = "ccc", digits = 2, caption = "Figure 1. Annual average of search square footage values.")

lhprism_yearly_stat <- lhprism %>%
    filter(!is.na(GROSS_TREAT_SQFT) | YEAR >= 2013) %>%
  group_by(YEAR) %>%
  summarize(across(c("GROSS_SQFT"), ~ mean(.x, na.rm = TRUE)))

            
kable(lhprism_yearly_stat, col.names = gsub("[_]", " ", names(lhprism_yearly_stat)), align = "ccc", digits = 2, caption = "Figure 2. Annual average of gross square footage values.")

lhprism_yearly_stat <- lhprism %>%
  filter(YEAR >= 2013) %>%
  group_by(YEAR) %>%
  summarize(across(c("SEARCH_SQFT", "GROSS_SQFT"), ~ sum(.x, na.rm = TRUE))) 
            
kable(lhprism_yearly_stat, col.names = gsub("[_]", " ", names(lhprism_yearly_stat)), align = "ccc", caption = "Figure 3. Annual total square footage values.")
```

```{r}
lhprism_plant_count <- lhprism %>%
  filter(!is.na(PLANT_COUNT)) %>%
  group_by(Species_Code) %>%
  summarize(Average_Plant_Count = mean(PLANT_COUNT, na.rm = TRUE), Total_Plant_Count = n(), SD_Plant_Count = sd(PLANT_COUNT, na.rm = TRUE)) %>%
  head()

lhprism_plant_count <- lhprism_plant_count[-1,]
kable(lhprism_plant_count, col.names = gsub("[_]", " ", names(lhprism_plant_count)), align = "lccc", digits = 2,  caption = "Figure 4. PLANT_COUNT values by species.")
```

# Data manipulation

### Finding unique population numbers of four target species.

#### HEMA17 (Giant hogweed). 163 rows.

```{r}
lhprism %>%
  filter(Species_Code == "HEMA17") %>%
  group_by(Species_Code,YEAR, VISIT_STATUS) %>% 
  summarize(GROSS_TREAT_SQFT, Treated_Pop = length(unique(Site))) %>%
  arrange(desc(YEAR)) %>%
  head()

lhprism %>%
  filter(Species_Code == "HEMA17") %>%
  group_by(Property_Pop, YEAR) %>% 
  summarize(Unique_Visit = length(unique(VISIT_STATUS))) %>%
  arrange(desc(Unique_Visit))

lhprism %>%
  filter(Property_Pop == "P020_DEC1850") %>%
  group_by(DAY, MONTH, YEAR) %>%
  arrange(desc(VISIT_DATE))
```

#### PUMO (Kudzu). 285 rows.

```{r}
lhprism %>%
  filter(Species_Code == "PUMO") %>%
  group_by(Species_Code,YEAR, VISIT_STATUS) %>% 
  summarize(GROSS_TREAT_SQFT, Treated_Pop = length(unique(Pop_Number))) %>%
  arrange(desc(YEAR)) %>%
  head()
```

#### CYSC4 (Scotch broom). 558 rows.

```{r}
lhprism %>%
  filter(Species_Code == "CYSC4") %>%
  group_by(Species_Code,YEAR, VISIT_STATUS) %>% 
  summarize(GROSS_TREAT_SQFT, Treated_Pop = length(unique(Pop_Number))) %>%
  arrange(desc(YEAR)) %>%
  head()
```

#### SAGL12 (Sticky sage). 199 rows.

```{r}
lhprism %>%
  filter(Species_Code == "SAGL12") %>%
  group_by(Species_Code,YEAR, VISIT_STATUS) %>% 
  summarize(GROSS_TREAT_SQFT, Treated_Pop = length(unique(Pop_Number))) %>%
  arrange(desc(YEAR)) %>%
  head()
```

#### HEMA17 (Giant hogweed)

```{r}
yearly_hema <- lhprism %>%
  filter(YEAR >= 2013, Species_Code == "HEMA17") %>%
  group_by(Species_Code, YEAR, VISIT_STATUS) %>%
  mutate(status_count = n())

ggplot(data = yearly_hema, aes(x = YEAR, fill = VISIT_STATUS)) + geom_bar(stat = "count")

ggplot(data = filter(lhprism, Species_Code == "HEMA17", Tot_Visits > 1), aes(x = VISIT_DATE, y = VISIT_STATUS, group = Property_Pop)) +
geom_line() + 
  facet_wrap(~Property_Pop)
```

#### PUMO (Kudzu).

```{r}
yearly_pumo <- lhprism %>%
  filter(Species_Code == "PUMO") %>%
  group_by(Species_Code, YEAR, VISIT_STATUS) %>%
  mutate(status_count = n())

ggplot(data = yearly_pumo, aes(x = YEAR, fill = VISIT_STATUS)) + geom_bar(stat = "count") 

# Large count of untreated in 2013 as predicated. Completely treated counts consistent until small decrease in later years. However, large count of Untreated in 2022. 

ggplot(data = filter(lhprism, Species_Code == "PUMO", Tot_Visits > 1), aes(x = VISIT_DATE, y = VISIT_STATUS, group = Property_Pop)) +
geom_line() + 
  facet_wrap(~Property_Pop)
```

#### CYSC4 (Scotch broom).

```{r}
yearly_scotch <- lhprism %>%
  filter(Species_Code == "CYSC4") %>%
  group_by(Species_Code, YEAR, VISIT_STATUS) %>%
  mutate(status_count = n())

ggplot(data = yearly_scotch, aes(x = YEAR, fill = VISIT_STATUS)) + geom_bar(stat = "count") 

# Large counts of Partially and Completely treated throughout years. Slight increase in Eradication in 2022. 

ggplot(data = filter(lhprism, Species_Code == "CYSC4", Tot_Visits > 1), aes(x = VISIT_DATE, y = VISIT_STATUS, group = Property_Pop)) +
geom_line() + 
  facet_wrap(~Property_Pop)
```

### SAGL12 (Sticky sage).

```{r}
yearly_sage <- lhprism %>%
  filter(YEAR >= 2013, Species_Code == "SAGL12") %>%
  group_by(Species_Code, YEAR, VISIT_STATUS) %>%
  mutate(status_count = n())

ggplot(data = yearly_sage, aes(x = YEAR, fill = VISIT_STATUS)) + geom_bar(stat = "count")

# Large count of partially treated in 2019 followed by its decrease and increase of Completely treated and 3/4 categories. 

ggplot(data = filter(lhprism, Species_Code == "SAGL12", Tot_Visits > 1), aes(x = VISIT_DATE, y = VISIT_STATUS, group = Property_Pop)) +
geom_line() + 
  facet_wrap(~Property_Pop)
```

### Verifying VISIT_DATE for target species.

#### SAGL12 (Sticky sage).

```{r}
lhprism %>% 
  filter(Species_Code == "SAGL12", Property_Pop == "ATNY_003") %>%
  select(Species_Code, VISIT_STATUS, VISIT_DATE, Property_Pop) %>%
  arrange(VISIT_DATE)

# ATNY_003 site visit dates and visit status are consistent with the previous facet_wrap plot. 

lhprism %>% 
  filter(Species_Code == "SAGL12", Property_Pop == "ATNY_005") %>%
  select(Species_Code, VISIT_STATUS, VISIT_DATE, Property_Pop) %>%
  arrange(VISIT_DATE)

#ATNY_005 - these dates look appropriate to the figure as well. 


lhprism %>% 
  filter(Species_Code == "SAGL12", Property_Pop == "P091_001") %>%
  select(Species_Code, VISIT_STATUS, VISIT_DATE, Property_Pop) %>%
  arrange(VISIT_DATE)

# P091_001 - these dates are consistent with the figure as the sharp incline is due to the last visit status date being No Plant 1st Yr. 

lhprism %>% 
  filter(Species_Code == "SAGL12", Property_Pop == "P091_009") %>%
  select(Species_Code, VISIT_STATUS, VISIT_DATE, Property_Pop) %>%
  arrange(VISIT_DATE)

# P091_009 - majority of these dates are in 2022 causing an accumulation of dates. Visit status fluctuates from various visit status categories. 


# Overall the dates seem consistent with the previously made figures. 
```

#### CYSC4 (Scotch broom)

```{r}
lhprism %>% 
  filter(Species_Code == "CYSC4", Property_Pop == "BEAR_006") %>%
  select(Species_Code, VISIT_STATUS, VISIT_DATE, Property_Pop) %>%
  arrange(VISIT_DATE)

# BEAR_006 - figure looks accurate to the visit dates.


lhprism %>% 
  filter(Species_Code == "CYSC4", Property_Pop == "HARR_003") %>%
  select(Species_Code, VISIT_STATUS, VISIT_DATE, Property_Pop) %>%
  arrange(VISIT_DATE)

# HARR_003 - figure looks accurate to visit dates, lots of back and forth from partially to completely treated and then back down to untreated. A plateau of completely treated status.

lhprism %>% 
  filter(Species_Code == "CYSC4", Property_Pop == "HARR_009") %>%
  select(Species_Code, VISIT_STATUS, VISIT_DATE, Property_Pop) %>%
  arrange(VISIT_DATE)

# HARR_009 - nearly the same trend as HARR_003. 
```

### Changed "No plants - 3 yrs" into "5-Eradicated". There are 20 cases of eradication. 1 case of CYSC4 (Scotch broom), 6 cases of PUMO (Kudzu), and 13 cases of HEMA17.

```{r}
lhprism %>%
  filter(VISIT_STATUS == "5- Eradicated") %>%
  select(VISIT_STATUS, Species_Code)

lhprism %>%
  count(VISIT_STATUS == "5- Eradicated", sort = TRUE)
```

### The sites where eradication occurred are P004, P005, P006, P018, and P029. P004, P005, P006, and P029 are located in Mahopac, NY, while P018 is in Fishkill, NY.

```{r}
lhprism_eradicated <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "HEMA17") %>%
  group_by(Species_Code, VISIT_STATUS) %>% 
  summarize(GROSS_TREAT_SQFT,Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head

kable(lhprism_eradicated, col.names = gsub("[_]", " ", names(lhprism_eradicated)), align = "llcc", format.args = list(big.mark = ","))
```

# Pull up method in relation to four target species.

### At first glance, there seems to be a large number of CYSC4, Scotch broom.

```{r}
lhprism %>%
  filter(!is.na(Species_Code), !is.na(PU_COUNT)) %>%
  group_by(Site, VISIT_STATUS, PU_COUNT, Species_Code) %>%
  summarize(Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(Site, VISIT_STATUS, PU_COUNT) %>%
  head
```

### Now only looking at Scotch broom cases under "pull up" treatment. There are 222 occurrences.

```{r}
scotch_pull_up <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "CYSC4") %>%
  group_by(Species_Code, VISIT_STATUS, PU_COUNT) %>%
  summarize(GROSS_TREAT_SQFT, Visit_Count=length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head

scotch_pull_up <- transform(scotch_pull_up, sum_vc = sum(Visit_Count))

# How should transform code look?

kable(scotch_pull_up, col.names = gsub("[_]", " ", names(scotch_pull_up)), align = "llccc", format.args = list(big.mark = ","))
```

### Just looking at SAGL12, Sticky sage. There are 26 total occurrences with NA PU_COUNT values included and 8 when such values are removed.

```{r}
sage_pull_up <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "SAGL12") %>%
  group_by(Species_Code, VISIT_STATUS, PU_COUNT) %>%
  summarize(GROSS_TREAT_SQFT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS)  %>%
  head

kable(sage_pull_up, col.names = gsub("[_]", " ", names(sage_pull_up)), align = "llccc", format.args = list(big.mark = ","))
```

### Looking at PUMO, Kudzu, plant under "pull up" method.

```{r}
kudzu_pull_up <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "PUMO") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, PU_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS)  %>%
  head

kable(kudzu_pull_up, col.names = gsub("[_]", " ", names(kudzu_pull_up)), align = "llccc", format.args = list(big.mark = ","))
```

### HEMA17, giant hogweed, under "pull up" method. 40 cases total without NA values.

```{r}
hogweed_pull_up <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "HEMA17") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, PU_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS)  %>%
  head

kable(hogweed_pull_up, col.names = gsub("[_]", " ", names(hogweed_pull_up)), align = "llccc", format.args = list(big.mark = ","))
```

# Foliar spray method in relation to four target species.

### There are 569 cases of recorded Foliar spray treatments.

```{r}
lhprism %>%
  group_by(Site, FO_COUNT, Species_Code) %>%
  summarize(Visit_Count = length(unique(VISIT_DATE)))  %>%
  head
```

### Looking at SAGL12, Sticky sage, cases only. 29 cases disregarding NA values.

```{r}
sage_fo <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "SAGL12") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, FO_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS)  %>%
  head

kable(sage_fo, col.names = gsub("[_]", " ", names(sage_fo)), align = "llccc", format.args = list(big.mark = ","))
```

### Looking at Giant hogweed under Foliar spray treatment. There are 22 recorded counts of the plant disregarding NA values of FO_COUNT.

```{r}
hogweed_fo <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "HEMA17") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, FO_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS)  %>%
  head

kable(hogweed_fo, col.names = gsub("[_]", " ", names(hogweed_fo)), align = "llccc", format.args = list(big.mark = ","))
```

### Looking at Scotch broom cases under Foliar spray treatment. There are 8 cases disregarding NA values.

```{r}
scotch_fo <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "CYSC4") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, FO_COUNT, Visit_Count = length(unique(VISIT_DATE)))  %>%
  head

kable(scotch_fo, col.names = gsub("[_]", " ", names(scotch_fo)), align = "llccc", format.args = list(big.mark = ","))
```

### There were no counts of FO_COUNT for HEMA17, giant hogweed.

# Cut only method in relation to four target species.

### There are 84 recorded cases of cut only method disregarding NA values.

```{r}
lhprism %>%
  filter(!is.na(CO_COUNT)) %>%
  group_by(Site, CO_COUNT, Species_Code) %>%
  summarize(Visit_Count = length(unique(VISIT_DATE)))  %>%
  head
```

### Looking at Scotch broom only, there are 24 total cases under Cut only treatment with 10 being NA values and 14 recorded values.

```{r}
scotch_co <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "CYSC4") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, CO_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS)  %>%
  head

kable(scotch_fo, col.names = gsub("[_]", " ", names(scotch_fo)), align = "llccc", format.args = list(big.mark = ","))
```

### Looking at SAGL12, Sticky sage, cases. There are 19 cases including NA values. Only one case with a recorded CO_COUNT value.

```{r}
sage_co <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "SAGL12") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, CO_COUNT, Visit_Count = length(unique(VISIT_DATE)))  %>%
  head

kable(sage_co, col.names = gsub("[_]", " ", names(sage_co)), align = "llccc", format.args = list(big.mark = ","))
```

### There are 14 cases of HEMA17 under the Cut only method. 12 out of the 14 cases under "2- Completely Treated" for VISIT_STATUS recorded a "0" in CO_COUNT.

```{r}
hogweed_co <- lhprism %>%
  filter(Species_Code == "HEMA17") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_SQFT, CO_COUNT, Visit_Count = length(unique(VISIT_DATE)))%>%
  arrange(VISIT_STATUS)  %>%
  head

kable(hogweed_co, col.names = gsub("[_]", " ", names(hogweed_co)), align = "llccc", format.args = list(big.mark = ","))
```

### Now just looking at PUMO, Kudzu, cases. There are 77 total cases including NA values but only one disregarding them.

```{r}
kudzu_co <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "PUMO") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, CO_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS)  %>%
  head

kable(kudzu_co, col.names = gsub("[_]", " ", names(kudzu_co)), align = "llccc", format.args = list(big.mark = ","))
```

# Cut and spray method in relation to four target species.

### There are 760 cases total and 383 without NA values.

```{r}
lhprism %>%
  filter(!is.na(CS_COUNT)) %>%
  group_by(Site, CS_COUNT, Species_Code) %>%
  summarize(Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(desc(CS_COUNT))
```

### Now only looking at CYSC4, Scotch broom, cases. There are 54 total and 52 disregarding NA values. Majority of treatment was done on BEAR and HARR sites which stand for Bear Mountain State Park and Harriman State Park respectively.

```{r}
scotch_cs <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "CYSC4") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, CS_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head()

kable(scotch_cs, col.names = gsub("[_]", " ", names(scotch_cs)), align = "llccc", format.args = list(big.mark = ","))
```

### Looking at HEMA17, Giant hogweed, cases. A count of 0 was recorded at 10 different sites. All sites were completely treated with cut and spray method. HEMA17 is documented as "5- Eradicated" as mentioned before. Successful method for HEMA17 control. However, the P006 site is the only site documented as Eradicated.

```{r}
hogweed_cs <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "HEMA17") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, CS_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head

kable(hogweed_cs, col.names = gsub("[_]", " ", names(hogweed_cs)), align = "llccc", format.args = list(big.mark = ","))
```

# Basal bark treatment in relation to four target species and AREL8.

### There are 705 total cases and 312 without NA "BB_COUNT" values. Large number of AREL8, Japanese angelica tree.

```{r}
lhprism %>%
  filter(!is.na(BB_COUNT)) %>%
  group_by(Site, BB_COUNT, Species_Code) %>%
  summarize(Visit_Count = length(unique(VISIT_DATE))) %>%
  head
```

### There are 175 cases of AREL8 under the basal bark treatment.

```{r}
angelica_bb <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "AREL8") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, BB_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head

kable(angelica_bb, col.names = gsub("[_]", " ", names(angelica_bb)), align = "llccc", format.args = list(big.mark = ","))
```

### There are no cases of CYSC4, Scotch Broom, under the basal bark treatment.

```{r}
scotch_bb <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "CYSC4") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, BB_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head

kable(scotch_bb, col.names = gsub("[_]", " ", names(scotch_bb)), align = "llccc", format.args = list(big.mark = ","))
```

### Again, there are no cases of SAGL12, Sticky sage.

```{r}
lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "SAGL12") %>%
  group_by(Site, VISIT_STATUS, BB_COUNT, Species_Code) %>%
  summarize(Visit_Count = length(unique(VISIT_DATE))) %>%
  head
```

### Again, there are no recorded cases of PUMO, Kudzu.

```{r}
lhprism %>%
  filter(!is.na(BB_COUNT), Species_Code == "PUMO") %>%
  group_by(Site, VISIT_STATUS, BB_COUNT, Species_Code) %>%
  summarize(Visit_Count = length(unique(VISIT_DATE)))
```

### There are 10 recorded cases of HEMA17, Giant hogweed. All recorded counts are "0" and all were completely treated. Potentially a good treatment.

```{r}
hogweed_bb <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "HEMA17") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, BB_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head

kable(hogweed_bb, col.names = gsub("[_]", " ", names(hogweed_bb)), align = "llccc", format.args = list(big.mark = ","))
```

### P006 is a site of Eradication.

### Hack and spray treatment in relation to four target species and AREL8.

### There are 428 cases total and 22 disregarding NA values.

```{r}
lhprism %>%
  filter(!is.na(HS_COUNT)) %>%
  group_by(Site, HS_COUNT, Species_Code) %>%
  summarize(Visit_Count = length(unique(VISIT_DATE))) %>%
  head
```

### There are 3 cases of CYSC4, Scotch Broom, under hack and spray treatment; however all 3 are NA values.

```{r}
scotch_hs <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "CYSC4") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, HS_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head()

kable(scotch_hs, col.names = gsub("[_]", " ", names(scotch_hs)), align = "llccc", format.args = list(big.mark = ","))
```

### There are 18 cases of SAGL12, Sticky sage; however, all are NA values.

```{r}
sage_hs <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "SAGL12") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, HS_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head()

kable(sage_hs, col.names = gsub("[_]", " ", names(sage_hs)), align = "llccc", format.args = list(big.mark = ","))
```

### There are 10 cases of HEMA17, Giant hogweed. All cases recorded "0" and all were completely treated. Effective method.

```{r}
hogweed_hs <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "HEMA17") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, HS_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head()

 kable(hogweed_hs, col.names = gsub("[_]", " ", names(hogweed_hs)), align = "llccc", format.args = list(big.mark = ","))
```

### There are 77 cases of PUMO, Kudzu; however, all HS_COUNT values are NA.

```{r}
kudzu_hs <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "PUMO") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, HS_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
  arrange(VISIT_STATUS) %>%
  head

kable(kudzu_hs, col.names = gsub("[_]", " ", names(kudzu_hs)), align = "llccc", format.args = list(big.mark = ","))
```

### There are 69 total cases of AREL8, Japanese angelica tree and 6 when disregarding NA values.

```{r}
angelica_hs <- lhprism %>%
  filter(!is.na(GROSS_TREAT_SQFT), Species_Code == "AREL8") %>%
  group_by(Species_Code, VISIT_STATUS) %>%
  summarize(GROSS_TREAT_SQFT, HS_COUNT, Visit_Count = length(unique(VISIT_DATE))) %>%
    arrange(VISIT_STATUS) %>%
  head
  
kable(angelica_hs, col.names = gsub("[_]", " ", names(angelica_hs)), align = "llccc", format.args = list(big.mark = ","))
```

# Analysis of HEMA17 Eradication

```{r}
# Begin by creating new data frame of HEMA17 with a binary variable of eradicated or not. 

hema <- lhprism %>% 
  filter(Species_Code == "HEMA17") %>%
  arrange(VISIT_DATE)

hema$Status <- ifelse(hema$VISIT_STATUS == "5- Eradicated", "Eradicated","Not_Eradicated")

hema$Status <- as.factor(hema$Status)

# Using the lag function to find the previous plant count of each visit. 
hema$Previous_PC <- lag(hema$PLANT_COUNT, n=1)

# Look at duration population has been followed
hema_summary <-
  hema %>%
  group_by(Property_Pop) %>%
  summarise(Duration = max(VISIT_DATE) - min(VISIT_DATE),
            Num_visits = n(),
            Treatment_visits = sum(!is.na(PERCENT_TREATED)),
            Eradicated = as.numeric(as.logical(sum(Status == "Eradicated"))))


# Simple logistic regression of HEMA17. 
hema_model <- glm(Status ~ Previous_PC, data = hema, family = binomial(link='logit'))
hema_model
summary(hema_model)


hema_summary_mod <- glm(Eradicated ~ Duration * Treatment_visits, data = hema_summary, family = binomial(link = "logit"))
summary(hema_summary_mod)


ggplot(hema_summary, aes(x = Treatment_visits, y = Eradicated)) + 
  geom_jitter(width = 0.1, height = 0.1) +
  geom_smooth(method = "glm", se = FALSE,
       method.args = list(family = "binomial"), linetype = "dashed")

ggplot(hema_summary, aes(x = Duration, y = Eradicated, color = Treatment_visits)) + 
  geom_jitter(width = 0.1, height = 0.1) +
  geom_smooth(method = "glm", se = FALSE,
       method.args = list(family = "binomial"), linetype = "dashed")
```

#### Finding the best estimate of population size at the start and end.

```{r}
# Creating a data frame of maximum PLANT_COUNT values. 
hema_max <- hema%>%
  filter(!is.na(PLANT_COUNT)) %>%
  group_by(Property_Pop, VISIT_DATE) %>%
  summarise(Max_PC = max(PLANT_COUNT)) %>%
  mutate(First_Date = min(VISIT_DATE)) %>%
  arrange(VISIT_DATE)

# Using ifelse() function to create a new column to indicate if Property_Pop cases record their maximum PLANT_COUNT values on initial visit date or not.
hema_max$Match <- ifelse(hema_max$VISIT_DATE == hema_max$First_Date, "Yes", "No")

ggplot(hema_max, aes(x = VISIT_DATE, y = Max_PC)) + geom_smooth()

ggplot(hema_max, aes(x = VISIT_DATE, y = Max_PC, group = Property_Pop)) +
geom_line() + 
  facet_wrap(~Property_Pop)

# Creating new data frame of only cases where the recorded maximum PLANT_COUNT value matches the inital visit date.
hema_initial <- hema_max %>%
  filter(Match == "Yes") %>%
  rename(Initial_PC = Max_PC)

ggplot(hema_initial, aes(x = VISIT_DATE, y = Initial_PC)) + geom_smooth()

# Testing if I could use a logistic regression on the binomial factor of "yes" and "no" for the matching dates. 
ggplot(hema_max, aes(x = Max_PC, y = Match, color = Property_Pop)) + 
  geom_jitter() +
  geom_smooth(method = "glm", se = FALSE,
       method.args = list(family = "binomial"), linetype = "dashed")
```
