---
title: "Project Report Outline"
format: html
editor: visual
bibliography: references.bib
---

## Executive summary

Project overview. Written by all participants.

```{r package_setup}
## Get packages
library(pacman)
pacman::p_load(here, tidyverse, janitor, lubridate, leaflet, readxl)

## Get the project level directory path
proj_path <- here()

## Define a color blind friendly palette
## These are from - http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/
# The palette with grey:
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
# The palette with black:
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

```{r data_import}
## Read in the data file
## In this code chunk, we will read in the data file. You may have to edit this call, depending on where you have saved your version of the Excel file.

lhprism_data <- read_csv(paste0(proj_path, "/data/LHPRISM-Data.csv"), 
    col_types = cols(`iMap Treat` = col_number(), 
        GROSS_TREAT_SQFT = col_number())) %>%
  clean_names()

## Edit the order of levels for some of the variables
lhprism_data <- lhprism_data %>%
  mutate(native = factor(native, levels = c("low", "medium", "high")),
         non_nat = factor(non_nat, levels = c("low", "medium", "high")))

## Create a new species code column
lhprism_data$species_code <- str_extract(lhprism_data$site_code, "(?<=_)[^_]+(?=_)") 
## Remove plant column
lhprism_data$plant <- NULL

## Create a new site column using the species_code column
lhprism_data$site <- str_extract(lhprism_data$site_code, "^[^_]+")

## Remove all rows that have NA for site
lhprism_data <- lhprism_data %>%
  filter(!is.na(site))

# Make a column for number of visits for each Site Code.
lhprism_data <- lhprism_data %>%
  group_by(site_code) %>%
  mutate(tot_visits = length(unique(visit_date)))


```

## Introduction

Ecosystems around the world face many threats — things like habitat loss, climate change, and pollution [@gentili2021]. One problem that people often overlook is the spread of invasive species [@usda2024]. In New York State, this isn't just a future worry. It's something that scientists, state agencies, and local communities deal with every day [@nysdec2024].

This report uses field data from the Lower Hudson Partnership for Regional Invasive Species Management (LHPRISM) to look at where invasive species are found in the Lower Hudson Valley and how they are being managed [@nysdec2024]. To understand why this data matters, it helps to know the bigger picture: what invasive species are, how New York State organizes its response, and what official plans guide decisions about which species to focus on.

### Background on Invasive Species

At its most basic, an invasive species is any organism — plant, animal, insect, fungus, or microbe — that establishes itself outside its native range and causes meaningful harm to the environment, economy, or human health [@mack2000; @usda2024]. The word "invasive" is doing a lot of work in that definition. Not every non-native species causes problems; in fact, most introduced species fail to survive at all in new environments [@mack2000]. They arrive without the adaptations needed to handle unfamiliar climates, predators, and food sources, and they quietly die out. The ones that do establish themselves and thrive are the exception — but when they do, the consequences can be severe and long-lasting [@gentili2021].

What makes certain species so successful in new environments often comes down to a lack of natural checks [@mack2000]. Back in their home range, they are kept in balance by predators, competitors, parasites, and other ecological forces that have co-evolved alongside them over thousands of years. When they land somewhere new, those controls disappear [@mack2000]. Native species in the invaded area have no evolutionary history with this newcomer and may not recognize it as a threat, compete with it effectively, or survive in its presence [@usda2024]. The invasive species, freed from its usual constraints, can reproduce rapidly, consume resources aggressively, and spread far beyond where it first appeared [@gentili2021].

The pathways through which invasive species move around the world are closely tied to human activity [@mack2000]. Global trade is one of the biggest drivers — cargo ships, shipping containers, and pallets of imported goods routinely carry hitchhiking organisms across continents and oceans [@usda2024]. The pet trade has introduced countless exotic animals into new regions, and when those animals escape or are deliberately released, some find conditions favorable enough to establish wild populations [@usda2024]. Ornamental horticulture has a long history of introducing plants that eventually escape garden boundaries and spread into natural areas [@mack2000]. Even recreational activities like hiking, boating, and fishing can accidentally transport seeds, larvae, or spores from one watershed to another [@nysdec2024].

The ecological damage caused by established invasive species takes many forms [@gentili2021]. Some are aggressive predators that decimate native prey populations before those populations have any chance to adapt [@mack2000]. Others are competitors that crowd out native plants or animals from food, space, and other resources they need to survive [@usda2024]. Over time, a successful invasion can fundamentally restructure an ecosystem — replacing native plant communities with monocultures, disrupting food webs, altering water cycles, and in the most severe cases, driving native species to local or even global extinction [@gentili2021; @mack2000]. Invasive species can also carry parasites and pathogens that spill over into native wildlife, compounding the damage furtherndefined \[@us

Beyond the ecological toll, invasive species impose enormous economic costs [@diagne2021]. Agricultural industries lose billions of dollars annually to crop damage from invasive pests and weeds [@diagne2021]. Forestry operations contend with invasive insects and diseases that kill commercially valuable trees [@mack2000]. Water infrastructure is disrupted by organisms like zebra mussels that clog intake pipes and alter aquatic chemistry [@usda2024]. And the cost of active management — detection, treatment, removal, and restoration — adds yet another layer to the financial burden [@diagne2021; @leung2002]. In densely populated and ecologically diverse states like New York, these costs are felt across sectors and communities [@nysdec2024].

New York is particularly exposed to invasive species pressure for several interconnected reasons [@nysdec2024]. It sits at the hub of one of the busiest transportation networks on the eastern seaboard, with major ports, airports, highways, and rail lines that move enormous volumes of goods and people [@nysdec2024]. Its landscape spans a remarkable range of ecosystems — coastal wetlands, deciduous forests, alpine zones, freshwater river systems — each of which represents a potential foothold for a different suite of invasive organisms [@nysdec2024]. And its long history of settlement, agriculture, and land use transformation has already altered many native habitats in ways that make them more vulnerable to invasion [@gentili2021]. Put all of that together, and it becomes clear why New York has invested significantly in building a statewide infrastructure for invasive species management [@nysdec2024].

### Background on NYS DEC Invasive Species Programs and the PRISM System

The New York State Department of Environmental Conservation (DEC) works with many groups — such as resource managers, non-profit organizations, industry workers, recreation groups, and other state agencies — to deal with invasive species [@nysdec2024]. To help with this work, New York created eight Partnerships for Regional Invasive Species Management (PRISMs) [@nysdec2024]. These PRISMs cover the entire state and help reduce the damage invasive species cause to New York's environment [@nysdec2016]. PRISMs were formed and funded under New York State law (Environmental Conservation Law 9-1705) [@nysdec2024].

PRISMs are groups of partners who work together to manage invasive species in different regions of New York [@nysdec2016]. They include people from federal and state agencies, environmental groups, scientists, industry, outdoor recreation, and local community members [@nysdec2024]. The DEC supports each PRISM through the Environmental Protection Fund, giving money to the organizations that host and coordinate them [@nysdec2024; @nysdec2016].

### PRISMs carry out many important tasks, including:

Planning invasive species management for their region [@nysdec2016]

Running programs that help prevent new invasive species [@nysdec2016]

Surveying and mapping where invasive species are found [@nysdec2024]

Finding new invasions early and responding quickly [@nysdec2016]

Carrying out control and removal projects [@nysdec2016]

Restoring and monitoring habitats after treatment [@nysdec2024]

Teaching people about invasive species and their impacts [@nysdec2024]

Bringing partners together to coordinate work [@nysdec2016]

Recruiting and training volunteers [@nysdec2024]

Supporting research through citizen science with the Invasive Species Research Institute [@nysdec2024]

Reporting invasive species sightings to iMapInvasives [@nysdec2024]

Providing education and outreach to communities [@nysdec2016]

Since 2013, the DEC has contracted organizations to run all eight PRISMs across the state [@nysdec2016].

### PRISMs deliver:

Regional planning for invasive species management [@nysdec2016]

Early detection and rapid response [@nysdec2016]

Eradication and control projects [@nysdec2024]

Education and outreach (with DEC-supported educators) [@nysdec2024]

Coordination among partners [@nysdec2016]

Volunteer training and involvement [@nysdec2024]

Citizen-science support for research [@nysdec2024]

The DEC is the main state agency responsible for environmental protection in New York, and invasive species management is part of its work [@nysdec2024]. Because invasive species spread across boundaries, New York created PRISMs so each region can coordinate its efforts [@nysdec2024].

The Lower Hudson PRISM (LHPRISM) covers the Lower Hudson Valley [@nysdec2024]. It uses trained staff and community volunteers who go into the field, record invasive species, help with treatment projects, and upload data to a shared statewide database [@nysdec2024]. This data is used for many environmental projects in the region [@finley2023].

### Information and Background on the NYS Invasive Species Lists

Not every non-native species that shows up in New York needs immediate action [@leung2002]. Deciding which species deserve attention — and how urgent that attention should be — requires a clear system for evaluating risk [@leung2002]. New York State uses two main systems to do this: a regulatory list managed by the Department of Environmental Conservation (DEC) and a tiered priority system created through collaboration between the PRISM network, state agencies, and the New York Natural Heritage Program (NYNHP) [@nysdec2024; @nynhp2024].

The regulatory list, created under the Environmental Conservation Law, sorts species into two groups based on how harmful they are and what restrictions are needed [@nysdec2024].

Prohibited species are considered so damaging, and so unnecessary for any legal use, that they cannot be possessed, sold, bought, transported, or introduced in New York at all [@nysdec2024]. There are no permits for these species — they are simply banned [@nysdec2024].

Regulated species are more complicated [@nysdec2024]. They may have some allowed uses, but there are still rules in place to prevent them from spreading [@nysdec2024].

Before a species is added to either category, experts review scientific information about its ecological impacts, how quickly it spreads, how feasible it is to control, and any economic or social considerations [@leung2002; @nysdec2024].

Alongside this legal framework, NYNHP created a tiering system to help PRISMs and other managers decide where to focus their limited time and resources [@nynhp2024]. Because New York has hundreds of invasive species, the tiering system uses a structured method that combines iMapInvasives data, invasiveness assessments, and expert input [@finley2023]. A species only receives a tier if it is considered to have high negative impacts [@nynhp2024]. Species that are present but not very harmful do not get tier assignments [@nynhp2024]. The tier definitions were finalized in 2017 through a statewide collaboration, giving all PRISMs a shared set of priorities [@nynhp2024; @finley2023].

These tiers can change over time [@nynhp2024]. They depend on accurate data, so new reports or successful management efforts can shift a species' tier [@finley2023]. This is why reporting new sightings to iMapInvasives is strongly encouraged — better data leads to better decisions [@nynhp2024]. For the species included in this report, their presence in the LHPRISM dataset means they have already been evaluated, found to be high-impact, and identified as priorities for management in the Lower Hudson region [@finley2023; @nysdec2024].

Understanding this background — the regulatory status, the tier level, and the management history of each species — is important for interpreting the data and understanding what it means for ongoing conservation work in this part of New York State [@nysdec2024; @nynhp2024].

## Region of study - Tanaka

The study area for this project encompasses the Lower Hudson region, as delineated by the Lower Hudson Partnership for Regional Invasive Species Management (LH PRISM). Established as one of eight regional invasive species management groups across New York State, LH PRISM coordinates collaborative efforts among government agencies, conservation organizations, and landowners to monitor and manage invasive species at a regional scale [@lhprism2024]. The region includes Westchester, Dutchess, Putnam, Rockland, Orange, and Ulster counties, extending southward into the five boroughs of New York City [@lhprism2024; @nysdec2024].

Geographically, the region is anchored by the Hudson River estuary, a highly dynamic tidal system in which freshwater inputs from inland watersheds interact with saltwater intrusion from the Atlantic Ocean. This estuarine gradient sustains a broad array of habitat types, including tidal wetlands, riparian corridors, freshwater marshes, and open aquatic environments, collectively rendering the Lower Hudson one of the most ecologically productive systems in the northeastern United States [@dec_aquatic2024]. The region also functions as a critical landscape transition zone, bridging the heavily urbanized New York City metropolitan area with the more rural and forested landscapes of the mid-Hudson Valley. This gradient of land use intensities is particularly significant in the context of biological invasions, as anthropogenic disturbance is well-documented as a primary driver of invasive species establishment and spread [@cornell2024; @nysdec2024].

### Climate

The Lower Hudson region is characterized by a humid temperate climate with four well-defined seasons. Summers are warm and humid, winters are cold with periodic snowfall, and precipitation is distributed relatively uniformly throughout the year, averaging approximately 40 to 50 inches annually [@noaa2024]. These climatic conditions support a wide diversity of native vegetation communities while simultaneously creating favorable thermal and moisture regimes for the establishment of non-native invasive species [@kunkel2013; @noaa2024]. Notably, climate projections for the northeastern United States indicate continued warming trends and increasing precipitation variability, conditions that are expected to further facilitate the range expansion of invasive species into previously unsuitable habitats [@kunkel2013].

### Ecology

The Lower Hudson region falls within the Eastern Temperate Forest biome and supports a mosaic of ecological communities, including mixed deciduous forests, freshwater wetlands, floodplain systems, and estuarine habitats [@nyheritage2021]. Dominant forest canopy species include oak (*Quercus* spp.), maple (*Acer* spp.), and hickory (*Carya* spp.), which provide structural complexity and support high levels of faunal biodiversity across the landscape [@nyheritage2021; @cornell2024]. The region's network of river systems and wetlands further contributes to its ecological significance, serving as corridors for both native wildlife and, unfortunately, the dispersal of invasive plant species [@dec_aquatic2024].

The Lower Hudson region presents an ideal study area for analyzing invasive species distribution patterns given its unique combination of ecological diversity, landscape heterogeneity, and anthropogenic pressure. The juxtaposition of urban and natural land cover types creates conditions highly conducive to invasive species introduction, as disturbance-adapted non-native plants readily exploit habitat edges, degraded areas, and human-modified landscapes [@lhprism2024; @nysdec2024]. The dataset employed in this study, compiled through field surveys conducted by the Invasives Strike Force under the coordination of LH PRISM, provides spatially explicit records of invasive plant populations across a broad range of property types within this region. Species identifications in the dataset are standardized using codes from the USDA PLANTS Database, ensuring taxonomic consistency across observation records [@usda2024]. Together, the ecological characteristics of the Lower Hudson region and the availability of a robust regional dataset make this an appropriate and well-suited geographic context for the analyses undertaken in this project.

## A map showing the representation of the Region of the Study is shown in the Spatial Analysis Section

**References**

Cornell University Department of Natural Resources and the Environment. (2024). *Conservation priorities in the watershed*. <https://hudson.dnr.cals.cornell.edu/natural-areas-biodiversity/conservation-priorities-watershed>

Kunkel, K. E., & others. (2013). *Regional climate trends and scenarios for the U.S. National Climate Assessment: Part 1. Climate of the Northeast U.S.* NOAA Technical Report NESDIS. <https://scenarios.globalchange.gov/sites/default/files/NOAA_NESDIS_Tech_Report_142-1-Climate_of_the_Northeast_US.pdf>

Lower Hudson Partnership for Regional Invasive Species Management. (2024). *About NYS PRISM network and the Lower Hudson*. <https://www.lhprism.org/about-us/>

National Centers for Environmental Information. (2024). *U.S. climate normals*. <https://www.ncei.noaa.gov/products/land-based-station/us-climate-normals>

New York Natural Heritage Program. (2021). *Ecological communities of New York State*. <https://guides.nynhp.org/ecological-communities/>

New York State Department of Environmental Conservation. (2024). *Aquatic habitats of the Hudson River Estuary*. <https://dec.ny.gov/nature/waterbodies/oceans-estuaries/hudson-river-estuary-program/aquatic-habitats>

New York State Department of Environmental Conservation. (2024). *Invasive species*. <https://dec.ny.gov/nature/animals-fish-plants/invasive-species>

USDA Natural Resources Conservation Service. (2024). *PLANTS database*. <https://plants.usda.gov>

## Overview of the LHPRISM data set

- Description of the data set and its contents

- An environmental monitoring dataset from the Lower Hudson Partnership for Regional Invasive Species Management focused on invasive species treatment in the Hudson Valley.

- What each record represents: A plot or site where invasive species were identified, treated, and monitored over time (3-year follow-up).

- Key variables: Location info ( site, participant) species present (abundance/cover) treatment type (chemical/method used) treatment timing (initial + follow-ups) monitoring outcomes (change in invasive presence over time) Missing/NULL values (HUGE PROBLEM WHAT TO SAY ABOUT MISSING DATA ITS EVERYWHERE)

- Description of data collection methods (will require communication with LHPRISM staff)

- Field surveys conducted at designated plots, identification of invasive species and estimation of their abundance, application of treatment, follow up monitoring intervals (yearly for 3 years)

- NEED to zoom with Brent (whether plots are randomly selected or targeted, this matters a lot, whether measurement methods are standardized)

- More questions: what defines a “plot”, [**what do NULL values represent**]{.underline} ,monitoring consistency? known biases (sites chosen because they were heavily bad)

- Summary statistics and figures about major components of the data set (e.g., number of records, number of species, etc.)

- Total number of rows, number of unique plots, number of invasive species tracked number of treatment types, time span

- Figures? Histogram of number of species, count of treatments applied

- [**Missing data visualizatio**n]{.underline}

- Should we put analyses of correlations and relationships among the variables in this section?

- No, this is what data do we have not what does it tell us or what can we glean, **MISSSING DATA**

- In regards to reaching out, Mimi and I emailed Brent some of these initial question listed here and asked to meet on zoom about more specific ones.

## Spatial distribution of the data

The Leaflet JavaScript library was used to create two maps for this dataset: one depicting the exact locations of each population that was treated and another displaying their treatment statuses. Map one displays the properties at which the invasive species in this project were observed.Users can interact with it by clicking on each data point to see the name of its corresponding property. Map two describes the extent to which the populations at each location were treated using a color coordination scheme. These maps provide a tidy visual aid for users by offering a straightforward presentation with a useful interface that allows users to learn more about any data point they choose.

Leaflet relies heavily on the pipe function to allow users to add layers to maps. To create the first map, users will load four packages: leaflet, readxl, tidyverse, and here. After loading these packages and reading in the data, the pipe operator is used with the filter function to eliminate NA values from the data. Then, the "\$" extraction operator is used to comb through the data, isolating points by latitude, longitude, and property name as separated by columns in Excel.

In the LHPRISM data set, there is a point whose longitude included an errant zero, putting it well out of range of the other data points. This error was corrected, incorporating the point back into the range of the others. Once the data is formatted appropriately, it can be displayed by creating a data frame with the isolated information related to location. Once this step is complete, the leaflet function itself is used to create the map object. The final step is to one again use the pipe operator, addTiles, and addMarkers functions to place the points on the map.

The second map uses a color gradient to classify points based on how far along the invasive species at these locations are in their treatment. The labels used are "Untreated," "Partially Treated," "Completely Treated," "No plants for 1 year," "No plants for 2 years," and "Eradicated." As with map one, the pipe operation is used with the addTiles and addCircleMarkers functions. These functions are used to assign different colors, along a spectrum, to pointes as determined by the status of their treatment. Untreated sites are grey, partially treated sites are orange-red, completely treated sites are yellow, sites with no plants for a year are yellow-green, sites with no plants for two years are green, and sites that have been eradicated of invasive plants are represented with lime green. This gradient provides a display that is easy for readers to understand and follow along with.

- Leaflet maps of populations:

```{r map_1}
# Patch to the Excel sheet
excel_file <- paste0(proj_path, "/data/Pace Data Project Invasives Strike Force Records.xlsx")

# Read in sheet 2 of the Excel spreadsheet (populations)
lhprism_pops <- read_excel(path = excel_file, sheet = 2)

# Remove NA values
lhprism_pops <- lhprism_pops %>%
  filter(!is.na(LONG))

# Isolate Latitude, Longitude, Population status and Property name
location_name <- lhprism_pops$PROPERTY_NAME
location_lat <- lhprism_pops$LAT
location_lng <- lhprism_pops$LONG

# Fixed data point that in the ocean due to incorrect data entry.
# Latitude of 40.0975937 changed to 40.975937 (Rye Nature Center)
location_lat[234] <- 40.975937

# Build data frame with latitude and longitude values
df1 <- data.frame(name = location_name,
                  lat = location_lat,
                  lng = location_lng)
# Create leaflet map object
m1 <- leaflet()

# Generate points for each population's coordinates
# Click on each point to see the Property Name
m1 %>%
  addTiles() %>%
  addMarkers(data = df1, popup = location_name)

```

Map 1: Locations of each treatment population. If viewing the interactive map, clicking on each point will bring up its associated property name (e.g. Rye Nature Center).

```{r map_2}
# Fix status text: ""1-Partially Treated" instead of "1- Partially Treated""
lhprism_pops$POP_STATUS[3] <- "1- Partially Treated"
lhprism_pops$POP_STATUS[7] <- "1- Partially Treated"
lhprism_pops$POP_STATUS[10] <- "1- Partially Treated"
lhprism_pops$POP_STATUS[15] <- "1- Partially Treated"

# Fix status text: ""2-Completely Treated" instead of "2- Completely Treated""
lhprism_pops$POP_STATUS[1] <- "2- Completely Treated"
lhprism_pops$POP_STATUS[4] <- "2- Completely Treated"
lhprism_pops$POP_STATUS[5] <- "2- Completely Treated"
lhprism_pops$POP_STATUS[6] <- "2- Completely Treated"

# Build new data frames for each status category
stat_0 <- filter(lhprism_pops, POP_STATUS == "0- Untreated")
stat_1 <- filter(lhprism_pops, POP_STATUS == "1- Partially Treated")
stat_2 <- filter(lhprism_pops, POP_STATUS == "2- Completely Treated")
stat_3 <- filter(lhprism_pops, POP_STATUS == "3- No Plants 1 Yr")
stat_4 <- filter(lhprism_pops, POP_STATUS == "4- No Plants 2 Yrs")
stat_5 <- filter(lhprism_pops, POP_STATUS == "5- Eradicated")

# Fixing the same incorrectly entered latitude value (Rye Nature Center)
stat_1$LAT[37] <- 40.975937

# Generate points colored by visit status
m2 <- leaflet()

m2 %>%
  addTiles() %>%
  addCircleMarkers(data = stat_0,
                   lng=~stat_0$LONG,
                   lat=~stat_0$LAT,
                   popup=~paste0("Site Name: ",stat_0$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_0$POP_STATUS),
                   color="grey",
                   radius = 4,
                   opacity = 0.7,
                   fillColor = "white",
                   group= "0- Untreated"
  ) %>% 
  addCircleMarkers(data = stat_1,
                   lng=~stat_1$LONG,
                   lat=~stat_1$LAT,
                   popup=~paste0("Site Name: ",stat_1$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_1$POP_STATUS),
                   color="orangered",
                   radius = 4,
                   opacity = 0.8,
                   fillColor = "white",
                   group= "1- Partially Treated"
  ) %>% 
  addCircleMarkers(data = stat_2,
                   lng=~stat_2$LONG,
                   lat=~stat_2$LAT,
                   popup=~paste0("Site Name: ",stat_2$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_2$POP_STATUS),
                   color="yellow",
                   radius = 4,
                   opacity = 0.6,
                   fillColor = "white",
                   group= "2- Completely Treated"
  ) %>% 
  addCircleMarkers(data = stat_3,
                   lng=~stat_3$LONG,
                   lat=~stat_3$LAT,
                   popup=~paste0("Site Name: ",stat_3$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_3$POP_STATUS),
                   color="yellowgreen",
                   radius = 4,
                   opacity = 0.8,
                   fillColor = "white",
                   group= "3- No Plants 1 Yr"
  ) %>% 
  addCircleMarkers(data = stat_4,
                   lng=~stat_4$LONG,
                   lat=~stat_4$LAT,
                   popup=~paste0("Site Name: ",stat_4$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_4$POP_STATUS),
                   color="green",
                   radius = 4,
                   opacity = 0.8,
                   fillColor = "white",
                   group= "4- No Plants 2 Yr"
  ) %>% 
  addCircleMarkers(data = stat_5,
                   lng=~stat_5$LONG,
                   lat=~stat_5$LAT,
                   popup=~paste0("Site Name: ",stat_5$PROPERTY_NAME,"<br>",
                                 "Status: ",stat_5$POP_STATUS),
                   color="limegreen",
                   radius = 4,
                   opacity = 0.9,
                   fillColor = "white",
                   group= "5- Eradicated"
  ) %>%
  addLegend(
    position = "topleft",
    color=c("grey", "orangered", "yellow", "yellowgreen", "green", "limegreen"),
    labels=c("Untreated","Partially Treated","Completely Treated","No plants for 1 year", "No plants for 2 years", "Eradicated"),
    opacity = 0.9
  )
```

Map 2: Locations of each treatment population, colored by status of treatment (e.g. "Untreated", "Partially Treated"). If viewing the interactive map, clicking on each point will bring up its associated property name and visit status.

## Analysis of time / treatment effort required to eradication

- how many times did they need to visit a plot before eradication: look at total number of visits, and population numbers

- Description of the results of this analysis

- How Many site codes have been Eradicated (5)

- This will be done on a subset of species

- How did we choose these species?

- these species were chosen in part because LHPRISM believes eradication is possible and for some of these species like Giant Hogweed, specific traits that are hazardous to humans were also taken into account.

- Brief overview of ecology of these particular species

- Description of the methods used in this analysis

- Description of the results of this analysis

- accelerated growth, previous exploratory analysis file - HEMA17 - Giant Hogweed - PUMO - Kudzu - CYSC4 - Scotch Broom - SAGL12 - Sticky Sage - CYSC4, SAGL12, LECU, PLEIO, POPE10, RULA, AREL8, HEMA17, ACAR10, VISI, COIN21, PUMO, ARIT, ILCR2, PHVI81, SYPA12, VIDI80, COKO2, PHVI81, SYPA12, AKQU, AREL8, SPJA, WISI, PHVI81, BRPA4, MISI, RAFIB, BRSY, HUJA, HEHE, FRAL4, WISTE - ALL EXTERMINATED

- Brief overview of ecology of these particular species (add known methods of treatment, previous recommendation)

- HEMA17 - Giant Hogweed (Heracleum mantegazzianum)

- native to Eastern Europe/Western Asia, highly invasive in NE USA

- thrives in rich, moist soil, prefers sunlight and tolerant of partial shade

- found in river banks, streams, floodplains, roadsides + trafficked areas

- biennial lifecycle, plants 2+ yrs old 10-15ft tall

- 1 plant can produce 20,000-50,000 seeds, spreading by water/wind and human activity

- forms dense monocultures, increases erosion risk due to dying back in winter, leaves soil exposed along riverbanks

- phototoxic sap - sever skin burns with sun exposure

- treatment recommendation: chemical treatment in early spring, mechanical removal for small infestations

- PUMO - Kudzu

- native to East Asia, widespread in SE USA and migrating N

- thrives on roadside, forest edges, warm and humid climates, prefers full sun, climbs to canopy

- perennial vine, grows up to 1ft/day in optimal conditions

- spreads vegetatively, rooting at nodes

- forms dense monocultures, causes tree death by shading and structural stress

- alters nutrient cycle by nitrogen fixation

- treatment complex, requires multi-modal approach over multiple years

- CYSC4 - Scotch Broom

- Native to Europe, widespread invasiveness in PNW, California, and East Coast USA

- prefers sandy soil and full sun, found on roadsides, logged forests, grassland

- perennial shrub, lives up to 10-15 yrs

- 1 plant produces 1,000s of seeds

- dense thickets, reduces habitat quality for native species, alters nutrient cycle by nitrogen fixation

- seeds viable for 30+ yrs in soil makes eradication difficult

- hand pulling for small infestations, chemical spray (glyphosate, triclopyr, basal bark treatment, stem cutting, foliar spray

- SAGL12 - Sticky Sage

- native to Europe and West Asia

- prefers moist, well-drained soil, found in woodlands, shaded slopes, forest edges

- perennial herb, blooms late summer

- low ecological impact, coexists with native species

- known treatment: handpulling for young pop

- Description of the methods used in this analysis

- change in plant count over consecutive visits

- visit_date \* visit_status_num

```{r status_time}
##created a table to see how long does it take each species to move from one status to the next, sorting by site
status_timing <- lhprism_data %>%
  group_by(species_code, site, visit_status_num) %>%
  summarise(first_date = min(visit_date), .groups = "drop") %>%
  arrange(species_code, site, visit_status_num)

##this does not account for status relapses (yet) as it only searches for the earliest date that each visit_status_num was recorded
```

- maybe see if there are any patterns when sorting by herbicide/treatment method?

- how many times did they need to visit a plot before eradication: look at total number of visits, and population numbers

```{r hema17_status_time}
##HEMA17 population by site and visit_status_num
hogweed_pop <- lhprism_data %>%
  filter(species_code == "HEMA17") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count, gross_treat_sqft)

#plotting visit_status_num across time
ggplot(data = hogweed_pop, aes(x = visit_date, y = visit_status_num, color = site)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~site) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
  theme(axis.text.y = element_text(size = 5)) +
  scale_y_reverse()
```

```{r pumo_status_time}
##PUMO population by site and visit_status_num
jknotweed_pop <- lhprism_data %>%
  filter(species_code == "PUMO") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count, gross_treat_sqft)

#PUMO visit_status_num over time
ggplot(data = jknotweed_pop, aes(x = visit_date, y = visit_status_num, color = site)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~site) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
  theme(axis.text.y = element_text(size = 5)) +
  scale_y_reverse()

##plant_count at each status level
##days/visits until new status, also days until relapse?

#select(site, visit_date, visit_status_num, plant_count)

```

```{r kudo_status_time}
##KUDO population by site and visit_status_num
jknotweed_pop <- lhprism_data %>%
  filter(species_code == "KUDO") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count)
```

```{r cycsc4_status_time}
##CYSC4 population by site and visit_status_num
scotchbroom_pop <- lhprism_data %>%
  filter(species_code == "CYSC4") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count)
```

```{r sagl12_status_time}
##SAGL12 population by site and visit_status_num
stickysage_pop <- lhprism_data %>%
  filter(species_code == "SAGL12") %>%
  arrange(site, desc(plant_count)) %>%
  select(site, visit_date, visit_status_num, plant_count)

#PUMO visit_status_num over time
ggplot(data = stickysage_pop, aes(x = visit_date, y = visit_status_num, color = site)) +
  geom_line() +
  theme_minimal() +
  facet_wrap(~site) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
  theme(axis.text.y = element_text(size = 5)) +
  scale_y_reverse()
```

- Description of the results of this analysis \`\`\`

- How Many site codes have been Eradicated (5)

```{r}
lhprism_treated <-
  lhprism_data %>%
  filter(grepl("5", visit_status_num)) %>%
  mutate(Eradicated = ifelse(visit_status_num == "5", 1, 0))

```

- Count of visits 0-5 status

```{r}
Visit_num_vs_site_code <-
  lhprism_data %>%
  group_by(visit_status_num, site_code) %>%
  count() %>%
  filter(!is.na(site_code)) %>%
  pivot_wider(names_from = site_code, values_from = n)
lhprism_treated
```

- visit_status_num \* visit_date, examining how long it takes to progress to the next status for 4 target species

- HEMA17

```{r}
##HEMA17 population by site and visit_status_num
hogweed_status <- lhprism_data %>%
filter(species_code == "HEMA17") %>%
arrange(site, desc(plant_count)) %>%
select(site, visit_date, visit_status_num, plant_count, gross_treat_sqft)

#plotting visit_status_num across time
ggplot(data = hogweed_status, aes(x = visit_date, y = visit_status_num, color = site)) +
geom_line() +
theme_minimal() +
facet_wrap(~site) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
theme(axis.text.y = element_text(size = 5)) +
scale_y_reverse()

##plant_count at each status level?
##days/visits until new status, also days until relapse?
```

- PUMO

```{r}
##PUMO population by site and visit_status_num
jknotweed_status <- lhprism_data %>%
filter(species_code == "PUMO") %>%
arrange(site, desc(plant_count)) %>%
select(site, visit_date, visit_status_num, plant_count, gross_treat_sqft)

#PUMO visit_status_num over time
ggplot(data = jknotweed_status, aes(x = visit_date, y = visit_status_num, color = site)) +
geom_line() +
theme_minimal() +
facet_wrap(~site) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
theme(axis.text.y = element_text(size = 5)) +
scale_y_reverse()
```

- CYSC4

```{r}
##CYSC4 population by site and visit_status_num
scotchbroom_status <- lhprism_data %>%
filter(species_code == "CYSC4") %>%
arrange(site, desc(plant_count)) %>%
select(site, visit_date, visit_status_num, plant_count)

#CYSC4 visit_status_num over time
ggplot(data = scotchbroom_status, aes(x = visit_date, y = visit_status_num, color = site)) +
geom_line() +
theme_minimal() +
facet_wrap(~site) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
theme(axis.text.y = element_text(size = 5)) +
scale_y_reverse()
```

- SAGL12

```{r}
##SAGL12 population by site and visit_status_num
stickysage_status <- lhprism_data %>%
filter(species_code == "SAGL12") %>%
arrange(site, desc(plant_count)) %>%
select(site, visit_date, visit_status_num, plant_count)

#PUMO visit_status_num over time
ggplot(data = stickysage_status, aes(x = visit_date, y = visit_status_num, color = site)) +
geom_line() +
theme_minimal() +
facet_wrap(~site) +
theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5)) +
theme(axis.text.y = element_text(size = 5)) +
scale_y_reverse()
```

- Description of the results of this analysis

- HEMA17

```{r}
```

- PUMO

```{r}
```

- CYSC4

```{r}
```

- SAGL12

```{r}
```

## Conclusions and recommendations
